Towards multiversal modal operators for homotopy type theory

### 2023-11-29 slides-hottest2023.lagda.tex
First chapter finished

@@

  %\subhead{\emph{A naive formalization attempt}}
  \emph{A naive formalization attempt:}

  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
  \pause
  \qquad\scalebox{2}{\dbend}
  \medskip

@@

\begin{frame}{Sequences depending on an environment}
  Let~$X$ be an hset such that there is no surjection~$\NN \twoheadrightarrow X$.
  Then the type~$\NN \to X$ misses the \hil{generic enumeration} of~$X$.
  \medskip
  \pause

  \emph{Idea:} Approximate a (fictitious) surjection~$\NN \twoheadrightarrow X$
  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. Over time, such an
  approximation can grow to
  \begin{enumerate}
    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
    \sigma\tau$, so it becomes \emph{more surjective}.
  \end{enumerate}
  \medskip
  \pause

  Formally, we have the type~$\nabla\underline{X}$ of stage-dependent values of~$X$:
\end{frame}

>>>>

\newcommand{\genalpha}{\mbox{$\hspace{0.12em}\shortmid\hspace{-0.62em}\alpha$}}

\usepackage{newunicodechar}

\newunicodechar{∇}{\ensuremath{\nabla}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{∷}{\ensuremath{\!\!\!}}
\newunicodechar{⧺}{\ensuremath{\!\!\!}}
\newunicodechar{₁}{\ensuremath{_1}}
\newunicodechar{₂}{\ensuremath{_2}}
\newunicodechar{∈}{\ensuremath{\in}}

@@

  \emph{A naive formalization attempt:}

  \textbf{Def.} $\displaystyle\mathsf{Well}_0(X,{\leq}) \defeq
    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
  \pause
  \qquad\scalebox{1.5}{\dbend}
  \medskip

@@

\begin{frame}{Sequences depending on an environment}
  Let~$X$ be an hset such that there is no surjection~$\NN \twoheadrightarrow X$.
  Then the type~$\NN \to X$ misses the \hil{generic enumeration}~$\genalpha$ of~$X$.
  \medskip
  \pause

  \emph{Idea:} Approximate (fictitious) surjections~$\NN \twoheadrightarrow X$
  by finite sequences~$\sigma = x_0 \ldots x_{n-1} \? X^*$. \\ Over time, such an
  approximation can grow to
  \begin{enumerate}
    \item one of $\sigma y$, where~$y \? X$, so it becomes \emph{more defined}; or
    \item one of~$\sigma \tau$, where~$\tau \? X^*$ such that~$a \in
    \sigma\tau$, so it becomes \emph{more surjective}.
  \end{enumerate}
  \medskip
  \pause

  For~$k : \NN$ and~$a : X$, the expression~``$\genalpha\,k = a$'' does not
  denote a proposition but rather a \hil{stage-dependent proposition}~$X^* \to
  \mathsf{Prop}$, namely $\lambda x_0 \ldots x_{n-1}.\ k < n \times x_k = a$.
  \medskip
  \pause

  Given a stage-dependent proposition~$P$, $\hil{$\nabla P$}\,\sigma$ expresses that no
  matter how~$\sigma$ evolves to a better approximation~$\tau$,
  eventually~$P\,\tau$ will hold:
  \AgdaHide{
  \begin{code}
    {-# OPTIONS --no-prop --no-import-sorts --allow-unsolved-metas #-}
    open import Agda.Primitive using (Set)
    Prop = Set
    postulate
      X : Set
      X* : Set
      [] : X*
      _∷_ : X* → X → X*
      _⧺_ : X* → X* → X*
      _∈_ : X → X* → Prop
    module Foo where
  \end{code}
  }
  {\small\begin{code}
      data ∇ (P : X* → Prop) (σ : X*) : Prop where
        now : P σ → ∇ P σ
        later₁ : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
        later₂ : (a : X) → ((τ : X*) → a ∈ (σ ⧺ τ) → ∇ P (σ ⧺ τ)) → ∇ P σ
  \end{code}}

  \vspace*{-0.8em}
  For instance, for~$P\,x_0 \ldots x_{n-1} \defeq \exists_{k : \NN} (k < n
  \times x_k = a)$, we have~$\nabla P \, \varepsilon$. \quad ``$\genalpha$ is surjective''
\end{frame}

\begin{frame}{Well quasiorders revisited}
  \emph{A naive formalization attempt:}

  \textbf{Def.} $\displaystyle\mathsf{Well}_0(X,{\leq}) \defeq
    \prod\limits_{\alpha \? \NN \to X} \, \Bigl\|\sum\limits_{i \? \NN} \sum\limits_{j \? \NN}
      i < j \times \alpha\,i \leq \alpha\,j\Bigr\|_{-1}$
  \pause
  \qquad\scalebox{1.0}{\dbend}
  \bigskip

  \emph{An inductive rephrasing:}

  \textbf{Def.} $\displaystyle\mathsf{Well}(X,{\leq}) \defeq
    \nabla \mathsf{Good} \, \varepsilon$, where
  {\vspace*{-0.7em}\small\begin{code}
    data ∇ (P : X* → Prop) (σ : X*) : Prop where
      now : P σ → ∇ P σ
      later : ((y : X) → ∇ P (σ ∷ y)) → ∇ P σ
  \end{code}}

  \vspace*{-0.8em}
  In other words: \emph{A quasiorder~$X$ is well iff the generic sequence~$\NN \to X$
  is good.}

  \begin{block}{}
    \justifying
    \textbf{Prop.} Let~$X$ and~$Y$ be well quasiorders.
    \good{Without} assuming~\bad{\textsc{lem}},~$X \times Y$ is well.
  \end{block}

  \dnote{We have~$\mathsf{Well}(X,{\leq}) \to
  \mathsf{Well}_0(X,{\leq})$, but the converse only holds in presence of \hil{bar
  induction}. \emph{How much stronger exactly is the inductive rephrasing?}}
\end{frame}
